---
title: "Module 1: Tabular Data"
subtitle: "Working with larger-than-RAM data using duckdbfs"
author: "ESPM 288"
format: html
---

## Introduction

In this module, we will explore high-performance workflows for tabular data. We will use `duckdbfs` to work with datasets that are larger than available RAM by leveraging DuckDB's streaming and remote file access capabilities.

## Case Study: Global Supply Chains

We will be working with [EXIOBASE 3.8.1](https://source.coop/youssef-harby/exiobase-3), a global Multi-Regional Input-Output (MRIO) database. This dataset tracks economic transactions between sectors and regions, along with their environmental impacts (emissions, resource use, etc.).

**Data description:**
- **Coverage**: 44 countries + 5 rest-of-world regions.
- **Timeframe**: 1995–2022.
- **Content**: Economic transactions (Z matrix), final demand (Y matrix), and environmental stressors (F matrix).
- **Format**: Cloud-optimized Parquet, partitioned by year and matrix type.

## Setup

```{r}


library(duckdbfs)
library(dplyr)


```

## Exercise 1: connecting to remote data

We can open the entire dataset without downloading it using `open_dataset()`. The data is hosted on Source Cooperative. The `**` pattern allows recursive scanning of the partitioned parquet files.

```{r}
# Remote S3 path to EXIOBASE 3 (Source Cooperative)

duckdbfs::duckdb_secrets(
    key = "",
    secret = "",
    endpoint = "s3.amazonaws.com",
    region = "us-west-2"
)
s3_url <- "s3://us-west-2.opendata.source.coop/youssef-harby/exiobase-3/4588235/parquet/**"

# Open the dataset lazily
exio <- open_dataset(s3_url)

# View the schema (column names and types) without reading data
glimpse(exio)
```

## Exercise 2: Efficient Filtering

The dataset is large. We should filter *before* collecting any data into R.

```{r}
exio |>
    filter(year == 2022, region == "US") |>
    head() |> # view the first 6 rows
    collect()
```

## Exercise 3: CO2 Production Data

Let's read in CO2 production data from the F_satellite matrix, which contains raw environmental stressors in physical units.

```{r}
# Read CO2 emissions data for 2022
co2_data <- exio |>
  filter(matrix == "F_satellite", 
         str_detect(stressor, "CO2")) |>
  collect()

# View the structure
glimpse(co2_data)

# Check unique CO2 stressors
co2_data |>
  distinct(stressor, unit) |>
  arrange(stressor)
```

> **Task**: Construct a query to find the top 5 sectors in the US by CO2 emissions in 2022. Remember to check the column names in `exio` to find the appropriate emissions flow.

The first thing I need to do is look at what unique stressors are in the f impacts matrix. This is important to me because I want to know if there are multiple stressors measured for CO2. For example, combustion, vs cement manufacturing. 

```{r}
distinct_units <- exio %>%
  filter(matrix == "F_impacts") %>%
  distinct(stressor) %>%
  collect()
print(distinct_units, n = 50) 
```

Now I know that there are in fact multiple ways this matrix measures CO2. Before I get into that, I want to know what the total emissions are for the listed parameters. This includes methane and others. 

```{r}
exio |>
  filter(matrix == "F_impacts", year == 2022, region == "US") |>
  group_by(sector) |>
  summarise(total_emissions = sum(value, na.rm = TRUE)) |>
  arrange(desc(total_emissions)) |>
   collect() |>
  print(n=5)
```

Methane was not a lucky guess. I already knew since I ran the code. Maybe I should stop eating beef! 

```{r}
for_pretty  %>%
  distinct(unit) %>%
  print(n = 50)
 
```

I am a little worried that even though we are working with a matrix that the different CO2 stressors are going to be measured in a different unit so lets check that out. 
```{r}
for_pretty <-exio |>
  filter(matrix == "F_impacts", year == 2022, region == "US", str_detect(stressor, "CO2")) |>
  group_by(sector) |>
  summarise(total_emissions = sum(value, na.rm = TRUE)) |>
  arrange(desc(total_emissions))|>
   collect() |>
 slice_head(n=5)
```

Looks Good enough to me! There are two unites. One Gg CO2 and another unit that I am assuming is the equivilant. I won't know though without more background. Now we know based on the code block above that Production of electricity by coal  was the top CO2 producer in the US in 2022. A little backgound shows that Gg is gigagrams. A unit used for measuing very large amounts of CO2. 

# Visualize 
Here I am going to start visulizing the data 
```{r}
library(ggplot2)
library(scales)

GG_pretty <- for_pretty |>
  ggplot(aes(x = reorder(sector, total_emissions), y = total_emissions, fill = sector)) +
  geom_col(show.legend = FALSE) +
  scale_y_continuous(labels = comma) +
  labs(
    title = "Top 5 US Sectors by CO2 Emissions (2022)",
    x = "Sector",
    y = "Total CO2 Emissions (Gg)"
  theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5),
    axis.text.x = element_text(angle = 0, vjust = 0.5),
    axis.text.y = element_text(face = "bold"),
    panel.grid.major.y = element_blank()
  )
GG_pretty
```